The Effects of weather on maize yield: New evidence from Kenya

### 2018-12-25 writing/draft3/draft18.tex
writing

The preferred specification has a log-linear functional form, therefore the estimates in table~\ref{MainEst} are not directly interpretable as marginal effects and their effects are multiplicative. The predicted marginal effects of all coefficients are depicted in figures~\ref{MarEff1} in this section and in figures~\ref{MarEff2} and~\ref{MarEff3} in the appendix. The units of all predictors in all three figures are multiples of their standard deviations since the predictors were standardised. For the values of the marginal effects smaller than one, the effect on yield is negative while marginal effects higher than one can be interpreted as a positive effect. If, for example, a marginal effect of an explanatory variable $x$ was $1.1$ when $x$ is equal to its one standard deviation, one should interpret this as follows: if $x$ is equal to its standard deviation, the dependent variable (in our case maize yield) is $1.1$ times higher than it would be if the explanatory variable $x$ was equal to its mean (which is zero in figures~\ref{MarEff1} in this section and in figures~\ref{MarEff2} and~\ref{MarEff3} in the appendix as the predictors were standardised).

@@

The estimates of the model of the whole sample are summarised in the first two columns of table~\ref{MainEst}. Consistently with our assumptions, seasonal precipitation and average temperature have significant effects on yield. As we expected, the effect of temperature is negative. The linear term of precipitation is positive while the squared term of precipitation is negative and as we can see in figure~\ref{MarEff1}(a), the plot of the marginal effects of precipitation is hill-shaped. That is, the effect of precipitation is negative and increasing if the amount of precipitation is below its mean and the marginal effect increases with the amount of precipitation until it reaches $1.403$ times its standard deviation. At this point, the effect starts to decrease with increasing precipitation and it becomes negative again once the amount of precipitation reaches $2.805$ times its standard deviation. Therefore, too  much rain is harmful for agricultural yield as well as not enough rain. This is intuitive mainly because extreme rainfall can cause flooding which is damaging for yield.

@@

To sum up, in the ASAL counties temperature is much more important than in the non-ASAL counties where temperature is not significant. On the other hand, precipitation is not significant in the ASAL counties while it is strongly significant in the non-ASAL counties. This is illustrated by figure~\ref{MarEff1}. The plot of the marginal effects of precipitation is almost flat for the ASAL counties (figure \ref{MarEff1}(b)) while is steep and hill-shaped for the non-ASAL counties similarly as in the case of the whole sample. As one can see in figures~\ref{MarEff1}(d) and~\ref{MarEff1}(f), the marginal effects of temperature are linear and decreasing for the case of ASAL and non-ASAL counties. Marginal effects of the remaining predictors of the models for the two subsamples are depicted in figures~\ref{MarEff2}(c)-(f) and~\ref{MarEff3}(c)-(f) in the appendix.

@@

>>>>

The preferred specification has a log-linear functional form, therefore the estimates in table~\ref{MainEst} are not directly interpretable as the marginal effects and the coefficients are multiplicative. The predicted marginal effects of all coefficients are depicted in figures~\ref{MarEff1} in this section and in figures~\ref{MarEff2} and~\ref{MarEff3} in the appendix. The units of all predictors in all three figures are multiples of their standard deviations since the predictors were standardised. For the values of the marginal effects smaller than one, the effect on yield is negative while marginal effects higher than one can be interpreted as a positive effect. If, for example, a marginal effect of an explanatory variable $x$ was $1.1$ when $x$ is equal to its one standard deviation, one should interpret this as follows: if $x$ is equal to its standard deviation, the dependent variable (in our case maize yield) is $1.1$ times higher than it would be if the explanatory variable $x$ was equal to its mean (which is zero in figures~\ref{MarEff1} in this section and in figures~\ref{MarEff2} and~\ref{MarEff3} in the appendix as the predictors were standardised).

@@

The estimates of the model of the whole sample are summarised in the first two columns of table~\ref{MainEst}. Consistently with our assumptions, seasonal precipitation and average temperature have significant effects on yield. As we expected, the effect of temperature is negative. The linear term of precipitation is positive while the squared term of precipitation is negative and as we can see in figure~\ref{MarEff1}(a), the plot of the marginal effect of precipitation is hill-shaped. That is, the effect of precipitation is negative and increasing if the amount of precipitation is below its mean and the marginal effect increases with the amount of precipitation until it reaches $1.403$ times its standard deviation. At this point, the effect starts to decrease with increasing precipitation and it becomes negative again once the amount of precipitation reaches $2.805$ times its standard deviation. Therefore, too  much rain is harmful for agricultural yield as well as not enough rain. This is intuitive mainly because extreme rainfall can cause flooding which is damaging for yield.

@@

To sum up, for the ASAL counties temperature is much more important than for the non-ASAL counties where temperature is not significant. On the other hand, precipitation is not significant for the ASAL counties while it is strongly significant for the non-ASAL counties. This is illustrated in figure~\ref{MarEff1}(c)-(f) in this section. The plot of the marginal effect of precipitation is almost flat for the ASAL counties (figure \ref{MarEff1}(c) in this section) while it is hill-shaped and steep for the non-ASAL counties (figure \ref{MarEff1}(e) in this section) similarly as in the case of the whole sample. As one can see in figures~\ref{MarEff1}(d) and~\ref{MarEff1}(f) in this section, the marginal effect of temperature is linear and decreasing for the case of ASAL and non-ASAL counties. Marginal effects of the remaining predictors are depicted in figures~\ref{MarEff2}(c)-(f) and~\ref{MarEff3}(c)-(f) in the appendix.

@@

Adjusted $R^2$ is not available for the linear mixed models. To get an approximation of the adjusted $R^2$, we estimated ordinary least squares (OLS) with the same explanatory variables which were included in our preferred linear mixed models as the fixed effects (see the list of variables in table~\ref{MainEst}). For each of the three samples (all counties, ASAL and non-ASAL) we estimated two variants of the OLS, one with the dummy variables for the counties and one without them. The values of the adjusted $R^2$ are summarasied in table~\ref{R2}.

\begin{threeparttable}
\singlespacing
\caption{Alternative to adjusted R2}\label{R2}
\begin{tabular}{@{}lll}
\\[-1em]
\br
\\[-1em]
&\multicolumn{2}{c}{\textbf{County level dummy variables:}}\\
\textbf{Sample:}&Yes&No\\
\textbf{All counties}& $0.507$&$0.307$\\
\textbf{ASAL}& $0.379$&$0.170$\\
\textbf{non-ASAL}&  $0.501$&$0.307$\\
\br
\end{tabular}

\end{threeparttable}

To check for presence of multicollinearity we calculated the variance inflation factors (VIF) of the explanatory variables. With only two exceptions, the VIF were below $2$ for all predictors in our final models. For CV of precipitation, the value of VIF was $2.172$ in the model of all counties, $2.147$ in the model for the subsample of the ASAL counties and $2.182$ in the model for the subsample of the non-ASAL counties. These values are still well below the threshold of problematic multicollinearity. The other exception was the seasonal precipitation as its VIF is $2.407$ in the model for the non-ASAL counties, probably as a result of the presence of the squared of precipitation. The VIF is below $2$ in the other two models and even the value $2.407$ is below the threshold of the problematic multicollinearity. Therefore, we can conclude that the models do not suffer from presence of multicollinearity.
